---
title: "annotate scRNA-seq reference"
author: "Lei Qin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
    keep_tex: no
    keep_md: true
    toc: yes
fontsize: 10pt
geometry: margin=1in
---

```{r style1, echo=FALSE, results='hide', message=FALSE, cache=FALSE}
library(knitr)
knitr::opts_chunk$set(
  fig.width=7, fig.height=7,
  fig.path='figure/',
  fig.align='center',
  out.width='90%',
  error=FALSE, prompt=TRUE, comment=NA,
  dpi=300, dev="png", dev.args=list(pointsize=15),
  cache=TRUE,
  cache.path="cache/",
  echo=TRUE,
  warning=FALSE,
  message=FALSE)

options(width=110, digits=3)
pdf.options(useDingbats = TRUE)
```

```{r setup, include=FALSE}
library(Seurat)
library(RColorBrewer)
library(magrittr)
library(dplyr)

cell_type_colors <- c(
  Tumor         = "#1F77B4",
  Myeloid       = "#AD8BC9",
  T             = "#D62728",
  Endothelial   = "#8C564B",
  Pericyte      = "#C49C94",
  Fibroblast    = "#FF7F0E",
  Adipocyte     = "#E377C2",
  B             = "#BCBD22",
  Plasma        = "#17BECF",
  Epithelial    = "#67BF5C",
  Lymph.vessel  = "#A2A2A2",
  unsure        = "black",
  B.Plasma      = "green"
)

ct_parent_groups <- list(
  Tumor        = c("Tumour"),
  Myeloid      = c("Monocyte", "Macrophage", "TAM", "cDC", "pDC", "Mast", "Neutrophil"),
  T            = c("CD4_T", "CD8_T", "Treg", "Ex_T", "NK"),
  Endothelial  = c("Endothelial"),
  Pericyte     = c("Pericyte"),
  Fibroblast   = c("Fibroblast", "iCAF", "myCAF", "apCAF"),
  Adipocyte    = c("Adipocyte"),
  B            = c("Naive_B", "Memory_B"),
  Plasma       = c("Plasma"),
  Epithelial   = c("Basal", "Luminal"),
  Lymph.vessel = c("Lymph.vessel")
)

source("plot_dotplot.R")
```

# MH0007

```{r MH0007-load}
seu_MH0007 <- readRDS("/vast/projects/Spatial/lei/Benchmarking/snRNA/MH0007/seu_MH0007.rds")
```

## Initial annotation — res 0.6

Uncertain clusters:

- **c2**: mixture of Plasma, Basal, Fibroblast, Tumour → assigned `unsure`
- **c12**: Pericyte / Basal / Fibroblast markers → assigned `Epithelial`

```{r MH0007-annot06, fig.width=8, fig.height=7}
seu_MH0007@meta.data %<>% mutate(cell_type0.6 = case_match(RNA_snn_res.0.6,
  c("0", "1", "3", "5", "11") ~ "Tumor",   # supported by inferCNV
  "4"                          ~ "B.Plasma",
  "6"                          ~ "Myeloid",
  "8"                          ~ "T",
  "10"                         ~ "Endothelial",
  "12"                         ~ "Epithelial",
  "13"                         ~ "Adipocyte",
  c("7", "9")                  ~ "Fibroblast",
  "2"                          ~ "unsure"
))

DimPlot(seu_MH0007, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

## Higher resolution — res 1.2

Reason: distinguish Pericyte from Endothelial.

```{r MH0007-cluster12, fig.width=8, fig.height=7}
seu_MH0007 <- FindClusters(seu_MH0007, resolution = 1.2, verbose = FALSE)
DimPlot(seu_MH0007, group.by = "RNA_snn_res.1.2",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r MH0007-dot12, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu_MH0007, "MH0007", group_by = "RNA_snn_res.1.2")
```

```{r MH0007-annot12a, fig.width=8, fig.height=7}
seu_MH0007@meta.data %<>% mutate(cell_type1.2 = case_when(
  RNA_snn_res.1.2 == "15" ~ "Pericyte",
  TRUE                     ~ cell_type0.6
))

DimPlot(seu_MH0007, group.by = "cell_type1.2", cols = cell_type_colors, label = TRUE)
```

Full re-annotation at res 1.2:

```{r MH0007-annot12b, fig.width=8, fig.height=7}
seu_MH0007@meta.data %<>% mutate(cell_type1.2 = case_match(RNA_snn_res.1.2,
  c("1", "2", "3", "5", "6", "7") ~ "Tumor",
  "10"                             ~ "T",
  "8"                              ~ "Myeloid",
  "4"                              ~ "Plasma",
  "12"                             ~ "Endothelial",
  "15"                             ~ "Pericyte",
  "14"                             ~ "Adipocyte",
  "13"                             ~ "Epithelial",
  c("11", "9")                     ~ "Fibroblast",
  "0"                              ~ "mixture"
))

DimPlot(seu_MH0007, group.by = "cell_type1.2", cols = cell_type_colors, label = TRUE)
```

## Subcluster B.Plasma (cluster 4 at res 1.2)

Attempting to separate B cells from Plasma. Result: subclustering failed to resolve a clean B population.

```{r MH0007-sub, fig.width=8, fig.height=7}
Idents(seu_MH0007) <- seu_MH0007$RNA_snn_res.1.2

seu_MH0007 <- FindSubCluster(
  seu_MH0007,
  cluster        = 4,
  graph.name     = "RNA_snn",
  resolution     = 0.8,
  subcluster.name = "subcluster"
)

DimPlot(seu_MH0007, group.by = "subcluster",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r MH0007-sub-dot, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu_MH0007, "MH0007", group_by = "subcluster")
```

## Save

```{r MH0007-save}
seu_MH0007$cell_type202605 <- seu_MH0007$cell_type1.2
saveRDS(seu_MH0007, "/vast/projects/Spatial/lei/Benchmarking/snRNA/MH0007/seu_MH0007.rds")
```

---

# MH0026

```{r MH0026-load}
seu <- readRDS("/vast/projects/Spatial/lei/Benchmarking/snRNA/MH0026/seu_MH0026.rds")
```

## Initial annotation — res 0.6

```{r MH0026-annot06, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_match(RNA_snn_res.0.6,
  c("0", "1", "2", "5", "10", "11") ~ "Tumor",   # supported by inferCNV
  "4"                                ~ "Fibroblast",
  "6"                                ~ "Myeloid",
  "7"                                ~ "T",
  "8"                                ~ "Endothelial",
  "9"                                ~ "Plasma",
  "13"                               ~ "Pericyte",
  "14"                               ~ "Adipocyte",
  c("3", "12")                       ~ "Epithelial" # used for Xenium validation
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

## Higher resolution — res 1.7

Reason: attempt to distinguish Lymph.vessel from Endothelial, and Plasma from B cells.

Result: Endothelial splits into 2 clusters but both co-express Endothelial and Lymph.vessel markers; B/Plasma not resolved.

```{r MH0026-cluster17, fig.width=8, fig.height=7}
seu <- FindClusters(seu, resolution = 1.7, verbose = FALSE)
DimPlot(seu, group.by = "RNA_snn_res.1.7",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

## Subcluster Myeloid (cluster 6 at res 0.6)

```{r MH0026-sub, fig.width=8, fig.height=7}
Idents(seu) <- seu$RNA_snn_res.0.6

seu <- FindSubCluster(
  seu,
  cluster         = 6,
  graph.name      = "RNA_snn",
  resolution      = 0.3,
  subcluster.name = "subcluster"
)

DimPlot(seu, group.by = "subcluster",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r MH0026-sub-dot, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "MH0026", group_by = "subcluster")
```

Subcluster 6_4 shows B cell markers → reassigned.

```{r MH0026-annot-sub, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_when(
  subcluster == "6_4" ~ "B",
  TRUE                ~ cell_type0.6
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

## Save

```{r MH0026-save}
seu$cell_type202605 <- seu$cell_type0.6
saveRDS(seu, "/vast/projects/Spatial/lei/Benchmarking/snRNA/MH0026/seu_MH0026.rds")
```

---

# ER_0114

```{r ER0114-load}
seu <- readRDS("/vast/projects/Spatial/lei/Benchmarking/scRNA/ER_0114/seu_ER_0114.rds")
```

## Initial annotation — res 0.6

Cluster notes:

- **c3, c12**: co-express Basal, Fibroblast, Pericyte markers → assigned `Epithelial`
- **c0, c1, c2, c4, c6, c10**: Tumour / Epithelial markers

```{r ER0114-cluster06, fig.width=8, fig.height=7}
seu <- FindClusters(seu, resolution = 0.6, verbose = FALSE)
DimPlot(seu, group.by = "RNA_snn_res.0.6",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r ER0114-annot06, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_match(RNA_snn_res.0.6,
  # confident
  "9"                              ~ "Fibroblast",
  c("8", "14")                     ~ "Myeloid",
  "13"                             ~ "Plasma",
  "11"                             ~ "Endothelial",
  "5"                              ~ "T",
  # less confident
  "7"                              ~ "Pericyte",
  c("0", "1", "2", "4", "6", "10") ~ "Tumor",
  c("3", "12")                     ~ "Epithelial"
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

## Subcluster Plasma (cluster 13 at res 0.6)

```{r ER0114-sub, fig.width=8, fig.height=7}
Idents(seu) <- seu$RNA_snn_res.0.6

seu <- FindSubCluster(
  seu,
  cluster         = 13,
  graph.name      = "RNA_snn",
  resolution      = 0.6,
  subcluster.name = "subcluster"
)

DimPlot(seu, group.by = "subcluster",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r ER0114-sub-dot, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "ER0114", group_by = "subcluster")
```

Subcluster 13_0 shows B cell markers → reassigned.

```{r ER0114-annot-sub, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_when(
  subcluster == "13_0" ~ "B",
  TRUE                  ~ cell_type0.6
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

## Save

```{r ER0114-save}
seu$cell_type202605 <- seu$cell_type0.6
saveRDS(seu, "/vast/projects/Spatial/lei/Benchmarking/scRNA/ER_0114/seu_ER_0114.rds")
```

---

# ER_0360

```{r ER0360-load}
seu <- readRDS("/vast/projects/Spatial/lei/Benchmarking/scRNA/ER_0360/seu_ER_0360.rds")
```

## Initial annotation — res 0.6

Uncertain clusters:

- **c14**: mixed Plasma / Myeloid markers → assigned `Myeloid`

```{r ER0360-annot06, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_match(RNA_snn_res.0.6,
  # confident
  c("2", "3", "8")   ~ "Tumor",
  c("0", "1", "5")   ~ "T",
  "4"                ~ "B",
  "7"                ~ "Myeloid",
  "9"                ~ "Endothelial",
  "10"               ~ "Pericyte",
  "12"               ~ "Fibroblast",
  # less confident
  "13"               ~ "Plasma",
  # not confident
  c("11", "6")       ~ "Epithelial",
  "14"               ~ "Myeloid"
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

```{r ER0360-dot06, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "ER0360", group_by = "RNA_snn_res.0.6")
```

## Save

```{r ER0360-save}
seu$cell_type202605 <- seu$cell_type0.6
saveRDS(seu, "/vast/projects/Spatial/lei/Benchmarking/scRNA/ER_0360/seu_ER_0360.rds")
```

---

# TN_0177

```{r TN0177-load}
seu <- readRDS("/vast/projects/Spatial/lei/Benchmarking/scRNA/TN_0177/seu_TN_0177.rds")
```

## Initial annotation — res 0.3

Uncertain clusters:

- **c9**: co-express Pericyte / Fibroblast markers
- **c13**: co-express Basal / Fibroblast / Pericyte markers
- **c14, c15**: co-express Myeloid / Fibroblast / T markers

```{r TN0177-annot03, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.3 = case_match(RNA_snn_res.0.3,
  # confident
  "6"              ~ "Tumor",
  c("1", "10")     ~ "Myeloid",
  c("0", "2", "3", "7") ~ "T",
  c("4", "8")      ~ "Plasma",
  "5"              ~ "B",
  "11"             ~ "Fibroblast",
  # less confident
  "12"             ~ "Endothelial",  # also shows Adipocyte markers
  "9"              ~ "Fibroblast",
  "13"             ~ "Epithelial",
  c("14", "15")    ~ "Myeloid"
))

DimPlot(seu, group.by = "cell_type0.3", cols = cell_type_colors, label = TRUE)
```

## Higher resolution — res 0.6

Reason: distinguish Pericyte from Fibroblast.

```{r TN0177-cluster06, fig.width=8, fig.height=7}
seu <- FindClusters(seu, resolution = 0.6, verbose = FALSE)
DimPlot(seu, group.by = "RNA_snn_res.0.6",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r TN0177-dot06, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "TN0177", group_by = "RNA_snn_res.0.6")
```

Cluster 18 at res 0.6 shows Pericyte markers:

```{r TN0177-annot06a, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_when(
  RNA_snn_res.0.6 == "18" ~ "Pericyte",
  TRUE                     ~ cell_type0.3
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

Full re-annotation at res 0.6:

```{r TN0177-annot06b, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.6 = case_match(RNA_snn_res.0.6,
  # confident
  "7"                          ~ "Tumor",
  c("1", "10", "12", "19")     ~ "Myeloid",
  c("0", "2", "3", "4", "8")  ~ "T",
  c("5", "15", "9")            ~ "Plasma",
  "6"                          ~ "B",
  c("11", "14", "17")          ~ "Fibroblast",
  "13"                         ~ "Endothelial",  # also shows Adipocyte markers
  "16"                         ~ "Epithelial",
  "18"                         ~ "Pericyte"
))

DimPlot(seu, group.by = "cell_type0.6", cols = cell_type_colors, label = TRUE)
```

## Higher resolution — res 2.5

Reason: look for Adipocytes. Result: cluster 12 (Endothelial/Adipocyte) did not split cleanly.

```{r TN0177-cluster25, fig.width=8, fig.height=7}
seu <- FindClusters(seu, resolution = 2.5, verbose = FALSE)
DimPlot(seu, group.by = "RNA_snn_res.2.5",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r TN0177-dot25, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "TN0177", group_by = "RNA_snn_res.2.5")
```

## Subcluster Endothelial (cluster 13 at res 0.6)

Reason: attempt to separate Adipocytes from Endothelial. Result: Adipocyte markers not resolved from Endothelial.

```{r TN0177-sub, fig.width=8, fig.height=7}
Idents(seu) <- seu$RNA_snn_res.0.6

seu <- FindSubCluster(
  seu,
  cluster         = 13,
  graph.name      = "RNA_snn",
  resolution      = 0.6,
  subcluster.name = "subcluster"
)

DimPlot(seu, group.by = "subcluster",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r TN0177-sub-dot, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "TN0177", group_by = "subcluster")
```

## Save

```{r TN0177-save}
seu$cell_type202605 <- seu$cell_type0.6
saveRDS(seu, "/vast/projects/Spatial/lei/Benchmarking/scRNA/TN_0177/seu_TN_0177.rds")
```

---

# TN_0554

```{r TN0554-load}
seu <- readRDS("/vast/projects/Spatial/lei/Benchmarking/scRNA/TN_0554/seu_TN_0554.rds")
```

## Initial annotation — res 0.3

Uncertain clusters:

- **c13**: Endothelial with Adipocyte co-expression

```{r TN0554-annot03, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.3 = case_match(RNA_snn_res.0.3,
  # confident
  c("1", "4")        ~ "Tumor",
  c("0", "10")       ~ "T",
  c("2", "5", "11")  ~ "Myeloid",
  "3"                ~ "B",
  c("6", "7")        ~ "Fibroblast",
  "8"                ~ "Plasma",
  "13"               ~ "Endothelial",  # also shows Adipocyte markers
  # less confident
  c("9", "12", "14") ~ "Epithelial"
))

DimPlot(seu, group.by = "cell_type0.3", cols = cell_type_colors, label = TRUE)
```

## Higher resolution — res 0.7

Reason: distinguish Pericyte from other stromal populations.

```{r TN0554-cluster07, fig.width=8, fig.height=7}
DimPlot(seu, group.by = "RNA_snn_res.0.7",
        reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5)
```

```{r TN0554-dot07, fig.width=14, fig.height=24, out.width='100%'}
plot_marker_dotplot(seu, "TN0554", group_by = "RNA_snn_res.0.7")
```

Cluster 17 at res 0.7 shows Pericyte markers:

```{r TN0554-annot07a, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.7 = case_when(
  RNA_snn_res.0.7 == "17" ~ "Pericyte",
  TRUE                     ~ cell_type0.3
))

DimPlot(seu, group.by = "cell_type0.7", cols = cell_type_colors, label = TRUE)
```

Full re-annotation at res 0.7:

```{r TN0554-annot07b, fig.width=8, fig.height=7}
seu@meta.data %<>% mutate(cell_type0.7 = case_match(RNA_snn_res.0.7,
  # confident
  c("1", "3", "12")          ~ "Tumor",
  c("2", "4", "9", "14", "20") ~ "T",
  c("5", "6", "7", "15")     ~ "Myeloid",
  "0"                         ~ "B",
  c("8", "11")                ~ "Fibroblast",
  "10"                        ~ "Plasma",
  "18"                        ~ "Endothelial",  # also shows Adipocyte markers
  "17"                        ~ "Pericyte",
  # less confident
  c("13", "16", "19")         ~ "Epithelial"
))

DimPlot(seu, group.by = "cell_type0.7", cols = cell_type_colors, label = TRUE)
```

## Save

```{r TN0554-save}
seu$cell_type202605 <- seu$cell_type0.7
saveRDS(seu, "/vast/projects/Spatial/lei/Benchmarking/scRNA/TN_0554/seu_TN_0554.rds")
```
