---
title: Human breast cancer sc/snRNA-seq analysis
author: "Lei Qin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
    keep_tex: no
    keep_md: true
    toc: no
fontsize: 10pt
geometry: margin=1in
---

```{r style1, echo=FALSE, results='hide', message=FALSE, cache=FALSE}
library(knitr)
# set global chunk options
knitr::opts_chunk$set(
  fig.width=7, fig.height=7,
  fig.path='figure/',
  fig.align='center',
  out.width='90%',
  #size="small",
  error=FALSE, prompt=TRUE, comment=NA,
  dpi=300, dev="png", dev.args=list(pointsize=15),
  cache=TRUE,
  cache.path="cache/",
  echo=TRUE, # set to FALSE to remove the code output
  warning=FALSE,
  message=FALSE)

options(width=110, digits=3)
pdf.options(useDingbats = TRUE)
```

```{r setup, include=FALSE}
library(Seurat)
library(tidyverse)
library(gprofiler2)
library(RColorBrewer)
library(scCustomize)
library(dplyr)
library(magrittr)
library(SingleR)
library(SingleCellExperiment)
library(pander)
library(tinytex)

source("/vast/projects/Spatial/lei/plotting_functions.R")
folder <- getwd()
```

## load data
```{r load data}
data = Read10X("/path/to/sample")
seu = CreateSeuratObject(data)
```

## Quality Control

The overview of transcript counts, gene numbers and mito gene percentage.
```{r QC, fig.width=9, fig.height=3, out.width='100%'}
seu$percent.mt = PercentageFeatureSet(seu, pattern = "^MT-")
VlnPlot(seu, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"))
```

```{r QC_plots, fig.width=9, fig.height=3, out.width='100%'}
seu@meta.data %>% as_tibble() %>%
  ggplot(aes(nFeature_RNA, ..density..)) +
  geom_density()+
  geom_vline(xintercept = 200) -> p1

seu@meta.data %>% as_tibble() %>%
  ggplot(aes(nCount_RNA, ..density..)) +
  geom_density()+
  geom_vline(xintercept = 10000) -> p2

seu@meta.data %>% as_tibble() %>%
  ggplot(aes(percent.mt, ..density..)) +
  geom_density()+
  geom_vline(xintercept = 10) -> p3

p1 | p2 | p3
```

```{r}
summary(seu$nFeature_RNA > 200 & seu$nCount_RNA < 10000 & seu$percent.mt < 10)
```

We remove cells with low detected genes, high total counts and high mitochondrial gene expression.
```{r}
seu = seu[, seu$nFeature_RNA > 200 & seu$nCount_RNA < 10000 & seu$percent.mt < 10]
```

## Data exploration

We log-normalize the data by library size, scale the data and perform dimensionality reduction by UMAP.

```{r process_seu}
seu = NormalizeData(seu)

seu = FindVariableFeatures(seu, selection.method = "vst", nfeatures = 2000)

seu = ScaleData(seu)

seu = RunPCA(seu, verbose = F, features = VariableFeatures(seu))
seu = RunUMAP(seu, dims = 1:25)

seu = FindNeighbors(seu, dims = 1:25)
```

We perform cell clustering and visualize the clusters on UMAP.

```{r CellClustering, message=FALSE, fig.height=7.5, fig.width=9, out.width='80%'}
seu = FindClusters(seu, resolution = 0.3)
DimPlot(seu, cols = colours, group.by = "seurat_clusters",label = T)
```

The heatmap shows the top 10 markers of each cluster.
```{r cluster_markers, fig.height=20, fig.width=14, out.width='100%'}
Idents(seu) <- seu$seurat_clusters
markers <- FindAllMarkers(seu, only.pos = T, logfc.threshold = 0.25, min.pct = 0.25)

markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(seu, features = top10$gene) + NoLegend()
```


## inferCNV

We infer copy number variation from expression data to identify cancer cells.

```{r, echo=FALSE}
infercnv_out_dir = paste0(folder, "/inferCNV_res")
```

The heatmap showing the cnv pattern:

```{r cnv_heatmap, echo = FALSE, out.width='100%', fig.align='center'}
infercnv_output_ht = paste0(infercnv_out_dir,"/","ComplexHeatmap_infercnv_with_sampleInfo_equal_width.png")
knitr::include_graphics(infercnv_output_ht)
```

The boxplot quantifying the cnv range for each group:

```{r cnv_boxplot, echo = FALSE, out.width='70%', fig.align='center'}
infercnv_output_boxplot = paste0(infercnv_out_dir,"/", "Boxplot_infercnv_instability_score",".png")
knitr::include_graphics(infercnv_output_boxplot)
```
